---
title: "BIOS 699 Project 3: Investigation of common chronic medication use and tumor sites in head and neck squamous cell carcinoma (HNSCC)"
author: "Mandy Ho"
date: "Analysis Complete: 3/20/2020"
header-includes:
- \usepackage{pdflscape}
- \newcommand{\blandscape}{\begin{landscape}}
- \newcommand{\elandscape}{\end{landscape}}
- \usepackage[english]{babel}
- \usepackage[utf8]{inputenc}
- \usepackage[absolute]{textpos}
- \usepackage{fancyhdr}
output: 
    pdf_document: default
---

```{r setup, include=FALSE, warning = FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(survival)
library(prodlim)
library(tidyverse)
library(survminer)
library(rms)
library(knitr)
library(missForest)
library(VIM)
require(cmprsk)
```


# I. Project Description

Head and neck squamous cell carcinoma (HNSCC) are cancer that mostly originate from lip, oral cavity, hypopharynx, oropharynx, nasopharynx or larynx. Prior research has identified several important associations between comorbidities medications with cancer outcome. However, there are lack of studies on how comorbidities drugs effect cancer outcome in rarer cancer such as HNSCC.

The goal for this study is to determine if common chronic medications (aspirin, metformin, statin, insulin) as well as different tumor sites (larynx, oropharynx, oral cavity) are associated with HNSCC patient’s survival outcomes.

# II. Study Design

Data were collected on N = 1642 HNSCC patients from the University of Michigan Specialized Program of Research Excellence (SPORE). Eligibility criteria including participants greater than 18 years of age, must be English speaker, no barriers which affect study compliance, incident of cancer must be no HNSCC diagnosed last 5 years and no previous treatment. Information were collected through patient interviews, pretreatment surveys, medical record extractions, annual chart reviews and follow up surveys, survival time and recurrence time were also collected throughout the study. Missing data imputation was conducted by Random Forest method.

# III. Statistical Analysis

Overall survival and progression-free survival were being assessed using Kaplan-Meier Estimators. Log-rank test was assessed to compare survival curves for covariates of interest that met proportional assumptions, while Wilcoxon test was assessed to compare survival curves that violated proportional assumptions.

Multivariate Cox Hazard models were built to evaluate the association between comorbidity drugs and tumor sites with HNSCC patient’s survival time. Additionally, Schoenfeld’s global test was assessed to test proportional hazard assumption in the Cox model, if the global proportional hazard assumption was violated (p < 0.05), risk factors that violated the proportional hazard assumption will be stratified in the model instead of including in the model as confounders.

```{r data_cleaning, echo=FALSE,  warning = FALSE, message=FALSE}
tdata = read.csv(file = "../HNSCC full data set 699 Project.csv", header = T)
# summary(tdata$how_old)
# create age category
tdata[tdata == ''] = NA
tdata['Age_group'] = ifelse(tdata$how_old <= 52, 1, 
                            ifelse( tdata$how_old > 52 & tdata$how_old <= 59, 2, 
                                   ifelse( tdata$how_old > 59 & tdata$how_old <= 68, 3,
                                           ifelse( tdata$how_old > 68, 4, NA))))
tdata$Age_group = factor(tdata$Age_group, levels = c(1,2,3,4), labels = c(1,2,3,4))

tdata['Larynx'] = ifelse(tdata$site == "Larynx" , 1, 0)
tdata['Oropharynx'] = ifelse(tdata$site == "Oropharynx" , 1, 0)
tdata['Oral Cavity'] = ifelse(tdata$site == "Oral Cavity" , 1, 0)
tdata['Other'] = ifelse(tdata$site == "Other (HP or other)" , 1, 0)

# tdata['drinking'] = ifelse(tdata$drinker == "current", 2,
#                            ifelse(tdata$drinker == "former drinker > 12 mos", 1,
#                                   ifelse(tdata$drinker == "never", 0, NA)))
# 
# tdata['smoking'] = ifelse(tdata$smoker == "current in past 12 mos", 2,
#                            ifelse(tdata$smoker == "former smoker > 12 mos",1,
#                                   ifelse(tdata$smoker == "never", 0, NA)))

# tdata['hpv'] = ifelse(tdata$hpvstat == "Positive", 1,
#                       ifelse(tdata$hpvstat == "Negative", 0,
#                              ifelse(tdata$hpvstat == "invalid/missing", NA, NA)))

tdata$drinker = factor(tdata$drinker, levels = c("never", "former drinker > 12 mos", "current"))
tdata$smoker = factor(tdata$smoker, levels = c("never", "former smoker > 12 mos", "current in past 12 mos"))
tdata$Aspirin = factor(tdata$Aspirin, levels = c(0,1))
tdata$metformin = factor(tdata$metformin, levels = c(0,1))
tdata$statin = factor(tdata$statin, levels = c(0,1))
tdata$Insulin = factor(tdata$Insulin, levels = c(0,1))

tdata$Larynx = factor(tdata$Larynx, levels = c(0,1))
tdata$Oropharynx = factor(tdata$Oropharynx, levels = c(0,1))
tdata$`Oral Cavity` = factor(tdata$`Oral Cavity`, levels = c(0,1))
tdata$Other = factor(tdata$Other, levels = c(0,1))
tdata$stage = factor(tdata$stage, levels = c(0,1,2,3,4))
```

## 1. Median follow-up time
```{r, echo = FALSE, warning = FALSE, message=FALSE}
# median follow up time
quantile(prodlim(Hist(stime, deathstatus) ~ 1, data = tdata,reverse=TRUE))
```


## 2. Missing data

### 2.1 Check missingness
```{r, echo=FALSE,  warning = FALSE, message=FALSE}
# check
aggr_plot = aggr(tdata, col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(tdata), cex.axis=0.7, gap=3, ylab=c("Histogram of missing data","Pattern"))

print(aggr_plot)
```

### 2.2 Missing data imputation with Random Forest (set trees = 200)
1. NRMSE (normalized mean squared error): represent error derived from imputing continuous values.     
2. PFC (proportion of falsely classified):  represent error derived from imputing categorical values.  
```{r, echo=FALSE,  warning = FALSE, message=FALSE}
# impute all data with RF, exclude ID, death and recurrence status
if (file.exists("data_imp_model.rds")){
  data.imp = readRDS("data_imp_model.rds") # load model
} else{
  data.imp = missForest(tdata %>% select(-c(idnum, deathstatus, recurstatus, stime, rtime)), 
                      ntree = 200)
  saveRDS(data.imp, "data_imp_model.rds") # save model
}

# imputation error
err = data.frame(model = "ntree=200", 
                 NRMSE = format(as.numeric(data.imp$OOBerror[1]), digits = 4), 
                 PFC = format(as.numeric(data.imp$OOBerror[2]), digits = 4))
kable(err, caption = "Imputation error using Random Forest ntree = 200")

# extract imputed data
new.imp = data.imp$ximp

# merge imputate data with ID, death and recurrence status
new.data = cbind(new.imp, tdata[,c("idnum", "deathstatus", "recurstatus", "stime", "rtime")])
```


## 3. Overall Survival

```{r os, fig.height = 6, out.width = c('50%', '50%'), echo=FALSE,  warning = FALSE, message=FALSE}
# all
os_fit = survfit(Surv(stime, deathstatus) ~ 1, data = new.data)
ggsurvplot(os_fit, data = new.data, surv.median.line = "hv", conf.int = TRUE, break.time.by = 20, risk.table = TRUE) + ggtitle("Overall Survival")

# by tumor site
os_fit_site = survfit(Surv(stime, deathstatus) ~ site, data = new.data)
ggsurvplot(os_fit_site, data = new.data, surv.median.line = "hv", pval = TRUE, pval.method = TRUE, conf.int = TRUE, break.time.by = 20, risk.table = TRUE, legend.title = "Tumor Site", legend = "top", legend.labs = c("Larynx", "Oral Cavity", "Orapharynx", "Other"),  font.main = 13, font.tickslab = 10, font.x =  11, font.y = 11) +
  ggtitle("Overall Survival by Tumor Site")

# by aspirin use
os_fit_asp = survfit(Surv(stime, deathstatus) ~ Aspirin, data = new.data)
ggsurvplot(os_fit_asp, data = new.data, surv.median.line = "hv", pval = TRUE, pval.method = TRUE, conf.int = TRUE,break.time.by = 20, risk.table = TRUE, legend.title = "Aspirin", legend = "top", legend.labs = c("No", "Yes"),  font.main = 13, font.tickslab = 10, font.x =  11, font.y = 11) +
  ggtitle("Overall Survival by Aspirin Use")

# by metformin use
os_fit_met = survfit(Surv(stime, deathstatus) ~ metformin, data = new.data)
ggsurvplot(os_fit_met, data = new.data, surv.median.line = "hv", pval = TRUE, pval.method = TRUE, conf.int = TRUE,break.time.by = 20, legend.title = "Metformin", risk.table = TRUE, legend = "top", legend.labs = c("No", "Yes"),  font.main = 13, font.tickslab = 10, font.x =  11, font.y = 11) +
  ggtitle("Overall Survival by Metformin Use")

# by statin use
os_fit_sta = survfit(Surv(stime, deathstatus) ~ statin, data = new.data)
ggsurvplot(os_fit_sta, data = new.data, surv.median.line = "hv", pval = TRUE, pval.method = TRUE, conf.int = TRUE,break.time.by = 20, risk.table = TRUE, legend.title = "Statin", legend = "top", legend.labs = c("No", "Yes"),  font.main = 13, font.tickslab = 10, font.x =  11, font.y = 11) +
  ggtitle("Overall Survival by Statin Use")

# by insulin use
os_fit_ins = survfit(Surv(stime, deathstatus) ~ Insulin, data = new.data)
ggsurvplot(os_fit_ins, data = new.data, surv.median.line = "hv", pval = TRUE, pval.method = TRUE, conf.int = TRUE,break.time.by = 20, risk.table = TRUE, legend.title = "Insulin", legend = "top", legend.labs = c("No", "Yes"),  font.main = 13, font.tickslab = 10, font.x =  11, font.y = 11) +
  ggtitle("Overall Survival by Insulin Use")

# +++++++++++++ BELOW NOT COVARIATE OF INTEREST +++++++++++ #
# by gender
# os_fit_sex = survfit(Surv(stime, deathstatus) ~ gender, data = new.data)
# ggsurvplot(os_fit_sex, data = new.data, surv.median.line = "hv", pval = TRUE, pval.method = TRUE, conf.int = TRUE,break.time.by = 20, risk.table = TRUE, legend.title = "Gender", legend = "top", legend.labs = c("Female", "Male"),  font.main = 13, font.tickslab = 10, font.x =  11, font.y = 11) +
#   ggtitle("Overall Survival by Gender")

# by age group
# os_fit_age = survfit(Surv(stime, deathstatus) ~ Age_group, data = new.data)
# ggsurvplot(os_fit_age, data = new.data, surv.median.line = "hv", pval = TRUE, pval.method = TRUE, conf.int = TRUE,break.time.by = 20, risk.table = TRUE, legend.title = "Age group", legend = "top", legend.labs = c("Age <= 52", "52 < Age <= 59", "59 < Age <= 68", "Age > 68"),  font.main = 13, font.tickslab = 10, font.x =  11, font.y = 11) +
#   ggtitle("Overall Survival by Age group")

# by drink
# os_fit_drink = survfit(Surv(stime, deathstatus) ~ drinker, data = new.data)
# ggsurvplot(os_fit_drink, data = new.data, surv.median.line = "hv", pval = TRUE, pval.method = TRUE, conf.int = TRUE,break.time.by = 20, risk.table = TRUE, legend.title = "Drinking status", legend = "top",  font.main = 13, font.tickslab = 10, font.x =  11, font.y = 11) +
#   ggtitle("Overall Survival by Drinking status")
```

## 4. Progression-Free Survival
```{r pfs, fig.height = 6, out.width = c('50%', '50%'), echo=FALSE,  warning = FALSE, message=FALSE}
# all
pfs_fit = survfit(Surv(rtime, recurstatus) ~ 1, data = new.data)
ggsurvplot(pfs_fit, data = new.data, surv.median.line = "hv", conf.int = TRUE,break.time.by = 20, risk.table = TRUE) + ggtitle("Progression-Free Survival")


# by tumor site
pfs_fit_site = survfit(Surv(rtime, recurstatus) ~ site, data = new.data)
ggsurvplot(pfs_fit_site, data = new.data, surv.median.line = "hv", pval = TRUE, pval.method = TRUE, conf.int = TRUE,break.time.by = 20, risk.table = TRUE, legend.title = "Tumor Site", legend = "top", legend.labs = c("Larynx", "Oral Cavity", "Orapharynx", "Other"),  font.main = 13, font.tickslab = 10, font.x =  11, font.y = 11) +
  ggtitle("Progression-Free Survival by Tumor Site")


# by aspirin use
pfs_fit_asp = survfit(Surv(rtime, recurstatus) ~ Aspirin, data = new.data)
ggsurvplot(pfs_fit_asp, data = new.data, surv.median.line = "hv", pval = TRUE, pval.method = TRUE, conf.int = TRUE,break.time.by = 20, risk.table = TRUE, legend.title = "Aspirin", legend = "top", legend.labs = c("No", "Yes"),  font.main = 13, font.tickslab = 10, font.x =  11, font.y = 11) +
  ggtitle("Progression-Free Survival by Aspirin Use")

# by metformin use
pfs_fit_met = survfit(Surv(rtime, recurstatus) ~ metformin, data = new.data)
ggsurvplot(os_fit_met, data = new.data, surv.median.line = "hv", pval = TRUE, pval.method = TRUE, conf.int = TRUE,break.time.by = 20, risk.table = TRUE, legend.title = "Metformin", legend = "top", legend.labs = c("No", "Yes"),  font.main = 13, font.tickslab = 10, font.x =  11, font.y = 11) +
  ggtitle("Progression-Free Survival by Metformin Use")

# by statin use
pfs_fit_sta = survfit(Surv(rtime, recurstatus) ~ statin, data = new.data)
ggsurvplot(os_fit_sta, data = new.data, surv.median.line = "hv", pval = TRUE, pval.method = TRUE, conf.int = TRUE, break.time.by = 20, risk.table = TRUE, legend.title = "Statin", legend = "top", legend.labs = c("No", "Yes"),  font.main = 13, font.tickslab = 10, font.x =  11, font.y = 11) +
  ggtitle("Progression-Free Survival by Statin Use") 

# by insulin use
pfs_fit_ins = survfit(Surv(rtime, recurstatus) ~ Insulin, data = new.data)
ggsurvplot(os_fit_ins, data = new.data, surv.median.line = "hv", pval = TRUE, pval.method = TRUE, conf.int = TRUE,break.time.by = 20, risk.table = TRUE, legend.title = "Insulin", legend = "top", legend.labs = c("No", "Yes"),  font.main = 13, font.tickslab = 10, font.x =  11, font.y = 11) +
  ggtitle("Progression-Free Survival by Insulin Use")
```


## 5. Multivariate Cox Model
* Note: Schoenfeld’s global test was assessed to test proportional hazard assumption in the Cox model, if the global proportional hazard assumption was violated (p < 0.05), risk factors that violated the proportional hazard assumption will be stratified in the model instead of including in the model as confounders.
* Outcome: overall survival

### 5.1. Time-independent *unadjusted* model: only covariate of interest included
* Covariate of interest: tumor sites (larynx, oropharynx, oral cavity), aspirin use, metformin use, statin use, insulin use
```{r cox, echo=FALSE,  warning = FALSE, message=FALSE}
cox_unadj = coxph(Surv(stime, deathstatus) ~ Aspirin + metformin + statin + Insulin + Larynx + Oropharynx + `Oral Cavity`, data = new.data)
summary(cox_unadj)
#extractAIC(cox_unadj)
ggforest(cox_unadj)

```

* AIC for the unadjusted model is `r extractAIC(cox_unadj)[2]`.

#### 5.1.1 Test for the proportional-hazards (PH) assumption
```{r, echo=FALSE,  warning = FALSE, message=FALSE}
cox.zph(cox_unadj)
```


### 5.2. Time-independent *adjusted* model: adjusting for confounder
* Covariate of interest: tumor sites (larynx, oropharynx, oral cavity), aspirin use, metformin use, statin use, insulin use
* Confounder: recurrence status, stage, smoker, drinker, age group, gender
```{r, echo=FALSE,  warning = FALSE, message=FALSE}
cox_adj = coxph(Surv(stime, deathstatus) ~  Aspirin + metformin + statin + Insulin + Larynx + Oropharynx + `Oral Cavity`+  factor(stage) + factor(smoker) + factor(drinker) + Age_group + gender, data = tdata)
summary(cox_adj)
# extractAIC(cox_adj)
```
* AIC for the adjusted model is `r extractAIC(cox_adj)[2]`.

#### 5.2.1 Test for the proportional-hazards (PH) assumption
```{r, echo=FALSE,  warning = FALSE, message=FALSE}
cox.zph(cox_adj)
```


\blandscape  
\section{5.2.2 Graphical diagnostic: scaled Schoenfeld residuals against the transformed time}   

```{r, fig.height=7, fig.width=12, ,echo=FALSE,  warning = FALSE, message=FALSE}
ggcoxzph(cox.zph(cox_adj), font.main = 13, font.tickslab = 10, font.x =  11, font.y = 11)
```

\elandscape  

### 5.3 Time-independent adjusted model with interaction
* Covariate of interest: tumor sites (larynx, oropharynx, oral cavity), aspirin use, metformin use, statin use, insulin use
* Confounder: recurrence status, stage, smoker, drinker, age group, gender
* Interaction: Aspirin * metformin,  Aspirin * statin,  Aspirin * Insulin,  metformin * statin,  metformin * Insulin, statin * Insulin   
```{r, echo=FALSE,  warning = FALSE, message=FALSE}
cox_adj_int = coxph(Surv(stime, deathstatus) ~ Aspirin*metformin + Aspirin*statin + Aspirin*Insulin + metformin*statin + metformin*Insulin + statin*Insulin + stage + smoker + drinker + Age_group + gender + Larynx + Oropharynx + `Oral Cavity`, data = new.data)

summary(cox_adj_int)
#extractAIC(cox_adj_int)
```
* AIC for adjusted model with interaction is `r extractAIC(cox_adj_int)[2]`. 

#### 5.3.1 Test for the proportional-hazards (PH) assumption
```{r, echo=FALSE,  warning = FALSE, message=FALSE}
cox.zph(cox_adj_int)
```



```{r time_varying, eval = F, echo=FALSE,  warning = FALSE, message=FALSE}
# time varying model: set recurrence time as time-varying parameter
tdata_long = tmerge(data1 = tdata %>% select(-rtime, -recurstatus), data2 = tdata, id = idnum, death = event(stime, deathstatus), recur = event(rtime, recurstatus))
attr(tdata_long, "tcount")
```


```{r, eval = F, echo=FALSE,  warning = FALSE, message=FALSE}
time_dep_cox = coxph(Surv(tstart, tstop, death) ~ recur + Aspirin + metformin + statin + Insulin + strata(site) + strata(`Age group`) + stage + smoker + drinker + strata(gender), data = tdata_long, ties = "breslow")
summary(time_dep_cox)
extractAIC(time_dep_cox)
```


```{r, eval = F, echo=FALSE,  warning = FALSE, message=FALSE}
time_dep_cox
```


## 6.Competing risk
```{r, echo=FALSE,  warning = FALSE, message=FALSE}
tsub = new.data[new.data$deathstatus == 1,]


tsub['cause'] = ifelse(tsub$cod == "Dead due to Malignancy", "Malignancy",
                       ifelse(tsub$cod == "Dead, cause unknown", "Unknown",
                              ifelse(tsub$cod == "Dead, other cause", "Other", NA)))

cr = cuminc(ftime = tsub$stime, fstatus = tsub$cause)

# Visualize
ggcompetingrisks(cr, palette = "Dark2",
                 legend = "top",
                 ggtheme = theme_bw())
```

